
```{r setup}
library(contextual)
library(simcontext)
```

# A single span of time

This is a very simple model (TASEP)
where there are two possible bases, `X` and `O`;
and `XO -> OX` at rate 1.
We will add a small amount of "noise" as well.

First we need to read in the basic model specification.
There are several steps to this:
```{r model}
model_json <- '
{
    "bases" : [ "X", "O" ],
    "initfreqs" : [ 0.5, 0.5 ],
    "mutpats" : [
        [ [ "XO", "OX" ] ],
        [ [ "X", "O" ], [ "O", "X" ] ]
    ],
    "mutrates" : [ 1, 0.05 ]
}
'
model_config <- read.config(json=model_json)
config <- parse.models(treeify.config(model_config,tlen=1))
```

Here is a short simulated sequence.
Mutated sites are shown in upper case;
unmutated sites that affected mutation rates are shown in lower case.
```{r simseq}
simseqs <- simseq.tree(200, config, count.trans=TRUE)
simseqs$tip
```

Now we will count T-mers.
These are not part of the model, so we specify these separately,
and then count up how many of each Tmer occurs in the sequence.
```{r opts}
opt <- list( longwin=5, shortwin=3, leftwin=1)

longpats <- getpatterns(opt$longwin,config$bases)
shortpats <- getpatterns(opt$shortwin,config$bases)

counts <- counttrans.list( list(longpats,shortpats), simseqs=simseqs, 
                leftwin=opt$leftwin, bases=config$bases )

```

## Generator and projection matrices

To get ready for inference we want to build some objects:
the generator matrix and the projection matrix.

The generator matrix:
```{r genmatrix}
genmatrix <- makegenmatrix(
                patlen=opt$longwin, 
                mutpats=config[["tip"]]$mutpats, 
                selpats=config[["tip"]]$selpats, 
                selfactors=config[["tip"]]$selfactors, 
                bases=config[["tip"]]$bases, 
                fixfn=config[["tip"]]$fixfn,
                Ne=config[["tip"]]$fixfn.params$Ne
            )
genmatrix@x <- update_x(G=genmatrix, mutrates=config[["tip"]]$mutrates, 
                      selcoef=config[["tip"]]$selcoef)
genmatrix
```

and the projection matrix:
```{r projemat}
projmatrix <- collapsepatmatrix( ipatterns=rownames(genmatrix), 
                                leftwin=leftwin(counts), 
                                fpatterns=colnames(counts) )
projmatrix
```

## A likelihood function

To compute the likelihood we need another set of counts and the corresponding projection matrix:

```{r more_setup}
shortpats_2 <- getpatterns(opt$shortwin-1, config$bases)
counts_2 <- counttrans.list( list(longpats,shortpats_2), simseqs=simseqs, 
                leftwin=opt$leftwin, bases=config$bases )
projmatrix_2 <- collapsepatmatrix( ipatterns=rownames(genmatrix), 
                                  leftwin=leftwin(counts_2), 
                                  fpatterns=colnames(counts_2) )
```


We can use these to construct a likelihood function
```{r likfun}
loglikfun <- function (x) {
    genmatrix@x <- update_x( G=genmatrix, mutrates=x, selcoef=c() )
    subtransmatrix <- computetransmatrix( genmatrix, projmatrix, tlen=1, time="fixed")
    subtransmatrix_2 <- computetransmatrix( genmatrix, projmatrix_2, tlen=1, time="fixed")
    # return POSITIVE log-likelihood
    num <- sum( counts@counts * log(subtransmatrix) )
    denom <- sum( counts_2@counts * log(subtransmatrix_2) )
    return(num-denom)
}

loglikfun(c(1,.05))
```




# A tree

Here is a simple configuration for a model with different mutation process on each branch.

```{r tree_model}
tree_model_json <- '
{
    "tree" : [ "(sp1 : 0.8, sp2 : 1.2) root;" ],
    "bases" : [ "X", "O" ],
    "initfreqs" : [ 0.5, 0.5 ],
    "sp1" : {
        "comment" : "Each X moves right.",
        "mutpats" : [
            [ [ "XO", "OX" ] ],
            [ [ "X", "O" ], [ "O", "X" ] ]
        ],
        "mutrates" : [ 1, .05 ]
    },
    "sp2" : {
        "comment" : "Each X moves left.",
        "mutpats" : [
            [ [ "OX", "XO" ] ],
            [ [ "X", "O" ], [ "O", "X" ] ]
        ],
        "mutrates" : [ 1, .05 ]
    }
}
'
tree_model_config <- read.config(json=tree_model_json)

tree_config <- treeify.config(tree_model_config)
tree_config <- parse.models(tree_config)
```
