---
title: "Checking the extension of an exponential time."
date: "`r date()`"
---

\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
```

Let $N$ be a Poisson process on $\R$,
let $L$ be a Poisson process on $\R^2$,
and let $T_t = \inf\{s>0 : L([0,s]\times[0,1/t])>0\}$.
The claim is that if we let
$$ Z(t) = N(T_t) $$
then this process has increments
$$
Z(t+u) - Z(t) \deq X_t + Y_t ,
$$
where $X_t$ is Bernoulli($t/(t+u)$)
and $Y_t$ is Geometric($(u+t)/(u+t+ut)$).

Here's the process:
```{r sim_process}
N <- cumsum(rexp(1e3))
min.t <- 1
max.x <- 1000
nL <- rpois(1,1/min.t*max.x) 
L <- cbind( runif(nL)*max.x, runif(nL)/min.t )
L <- L[order(L[,1],decreasing=FALSE),]
Tt <- function (t,L) {
    useL <- outer(L[,2],1/t,"<")
    L[ apply(useL,2,function(u)min(which(u))), 1 ]
}
Z <- function (t,L) {
    colSums(outer(N,Tt(t,L),"<"))
}
tt <- seq(min.t,500,length.out=1000)
layout(t(1:2))
plot( tt, Tt(tt,L), type='l', main="T(t)" )
plot( tt, Z(tt,L), type='l', main="Z(t)" )
```

Clearly, $Z(t+u)=Z(t)$ if the $y$-coordinate of the point of $L$
whose $x$-coordinate is $T_t$ is less than $1/(t+u)$;
this occurs with probability $t/(t+u)$.
If this is not the case,
then $T_t$ increments by an Exponential($1/(t+u)$) time,
which increments $Z$ by a Geometric($1/(1+(t+u)^{-1})=(t+u)/(1+t+u)$) number,
so we should have
$$ Z(t+u)-Z(t) \deq \text{Geom}\left(\frac{t+u}{1+t+u}\right) . $$


Increments of $T(t)$
------------------

We claim that
$$
T(t+u) \deq T(t) + \sum_{k=0}^M T_k,
$$
where the $T_k$ are independent copies of $T(t)$
and $M$ is Geometric($t/(t+u)$).

Here is the predicted distribution:
```{r geoexponential}
rgeoexp <- function (n,t,dt) {
    M <- rgeom(n,t/(t+dt))
    dT <- numeric(n)
    while ( any(M>0) ) {
        dT[M>0] <- dT[M>0] + rexp(n,1/t)[M>0]
        M <- M-1
    }
    return(dT)
}
rgs <- rgeoexp(1e4,t=10,dt=5)
mean(rgs==0)
```


And, the simulated distribution:
```{r sim_dT}
sim_dT <- function (t,dt) {
    N <- cumsum( rexp(5*(t+dt)) )
    max.x <- 100*t
    nL <- rpois(1,max.x/t)
    L <- cbind( runif(nL)*max.x, runif(nL)/t )
    L <- L[order(L[,1],decreasing=FALSE),]
    Tt(t+dt,L)-Tt(t,L)
}
sim.rgs <- replicate(1e4,sim_dT(t=10,dt=5))
mean(sim.rgs==0)
```

And the comparison for nonzero probabilities:
```{r compare_dT}
breaks <- seq(0,max(rgs,sim.rgs),length.out=50)
hist(rgs[rgs>0],breaks=breaks,freq=FALSE,col=adjustcolor("black",0.5))
hist(sim.rgs[sim.rgs>0],breaks=breaks,freq=FALSE,col=adjustcolor("red",0.5),add=TRUE)
```


Increments of $Z(t)$
------------------

Here's the predicted distribution:
```{r bernougeo}
rbernougeo <- function (n,t,dt) {
    ifelse( rbinom(n,1,t/(t+dt))>0, 0, rgeom(n,1/(1+t+dt)) )
}
bgs <- rbernougeo(1e6,t=10,dt=5)
mean(bgs==0)
```

And, the simulated distribution:
```{r sim_distrn}
sim_ppp <- function (t,dt) {
    N <- cumsum( rexp(5*(t+dt)) )
    max.x <- 100*t
    nL <- rpois(1,max.x/t)
    L <- cbind( runif(nL)*max.x, runif(nL)/t )
    L <- L[order(L[,1],decreasing=FALSE),]
    Z(t+dt,L)-Z(t,L)
}
sim.bgs <- replicate(1e4,sim_ppp(t=10,dt=5))
mean(sim.bgs==0)
```


```{r compare_Z}
breaks <- seq(0,max(bgs,sim.bgs,length.out=30))
hist(bgs[bgs>0],breaks=breaks,freq=FALSE,col=adjustcolor("black",0.5))
hist(sim.bgs[sim.bgs>0],breaks=breaks,freq=FALSE,col=adjustcolor("red",0.5),add=TRUE)
```
